---
title: "Architecture of the DuckDBArray Package"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r format(Sys.time(), '%B %d, %Y')`"
package: DuckDBArray
vignette: >
  %\VignetteIndexEntry{Architecture of the DuckDBArray Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 4
---

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

The `r Biocpkg("DuckDBArray")` package provides DuckDB-backed implementations
of `r Biocpkg("DelayedArray")` classes for efficient, out-of-memory operations
on large array and matrix datasets. It builds upon the `r Biocpkg("DuckDBDataFrame")`
foundation to deliver array-specific functionality with optimized matrixStats
methods and sparse array support.

## Position in the Bioconductor Ecosystem

`r Biocpkg("DuckDBArray")` sits in the middle layer of the DuckDB-backed
Bioconductor stack:

```
DuckDBDataFrame (foundation: tabular data + SQL)
    ↓
DuckDBArray (arrays/matrices + DelayedArray integration)
    ↓
BiocDuckDB (high-level I/O + analysis workflows)
```

| Class | Extends | Purpose |
|-------|---------|---------|
| *DuckDBArraySeed* | *Array* | Seed class for DelayedArray backend |
| *DuckDBArray* | *DelayedArray* | N-dimensional array |
| *DuckDBMatrix* | *DelayedMatrix* | 2D matrix specialization |

The package is designed for both **infrastructure developers** implementing
array-backed analyses and **end users** working with large genomic datasets.

## Installation

```{r install, eval=FALSE}
if (!require("BiocManager"))
    install.packages("BiocManager")
BiocManager::install("DuckDBArray")
```

```{r load-package}
library(DuckDBArray)
library(DuckDBDataFrame)  # For examples
```


# Core Design Principles

## The DelayedArray Backend Contract

`r Biocpkg("DuckDBArray")` implements a DelayedArray backend by providing:

1. **Seed class** (*DuckDBArraySeed*): Wraps a *DuckDBTable* from 
   `r Biocpkg("DuckDBDataFrame")`
2. **extract_array()** method: Materializes array blocks on demand
3. **extract_sparse_array()** method: Efficient sparse block extraction
4. **High-level classes**: *DuckDBArray* and *DuckDBMatrix* for user-facing API

This architecture enables:
- **Lazy evaluation**: Operations build a tree without immediate execution
- **Block processing**: Efficient chunked computation
- **Sparse optimization**: Minimal memory for sparse data
- **SQL translation**: Array operations become SQL aggregations


## Sparse Data Representation

Sparse arrays are naturally represented in Parquet because only non-zero
entries are stored. A sparse matrix with 1% density stores only 1% of the
full matrix size:

```
Dense: 10,000 × 10,000 = 100M entries
Sparse (1%): 1M non-zero entries in Parquet
```

DuckDB's columnar engine efficiently scans these non-zero entries.


## Grid Partitioning

For very large arrays, `r Biocpkg("DuckDBArray")` supports grid partitioning:

- Arrays are split into regular grid blocks
- Each block stored as a separate Parquet partition
- Dimension tables map indices to grid groups
- Operations process only relevant partitions

This enables parallel processing and efficient subsetting.


# The DuckDBArraySeed Class

## Class Structure

*DuckDBArraySeed* is the seed class that implements the DelayedArray backend:

```{r DuckDBArraySeed-class}
showClass("DuckDBArraySeed")
```

### Slots

**table** (*DuckDBTable*)
: A single-column DuckDBTable with `nkey(table) >= 1`. The keycols define
  the array dimensions, and the single datacol contains the array values.

**fill** (`atomic`)
: The fill value for missing entries (typically 0 for sparse arrays).

**drop** (`logical`)
: Whether to drop dimensions of length 1 (for subsetting operations).


## The Seed Contract

*DuckDBArraySeed* implements the required seed methods:

### dim() and dimnames()

```{r DuckDBArraySeed-dim}
# Create a simple 3D array seed
array_df <- expand.grid(
    i = 1:4,
    j = 1:3,
    k = 1:2
)
array_df$value <- rpois(nrow(array_df), lambda = 2)

array_path <- tempfile(fileext = ".parquet")
arrow::write_parquet(array_df, array_path)

seed <- DuckDBArraySeed(array_path,
                        datacol = "value",
                        keycols = list(i = 1:4, j = 1:3, k = 1:2))

# Dimensions from keycols
dim(seed)

# Dimension names
dimnames(seed)
```

### extract_array()

This is the core method that materializes array blocks:

```{r DuckDBArraySeed-extract}
# Extract a block
extract_array(seed, list(1:2, 1:2, 1))
```

The implementation:
1. Subsets the DuckDBTable to the requested indices
2. Calls `as.data.frame()` to materialize the data
3. Maps the results into an array with proper dimensions
4. Fills missing entries with the fill value


### extract_sparse_array()

For sparse data, this method returns a `COO_SparseArray`:

```{r DuckDBArraySeed-sparse}
# Create sparse data (only non-zero entries)
sparse_df <- data.frame(
    i = c(1, 1, 2, 3, 4),
    j = c(1, 3, 2, 1, 3),
    k = c(1, 1, 2, 2, 1),
    value = c(5, 3, 8, 2, 7)
)

sparse_path <- tempfile(fileext = ".parquet")
arrow::write_parquet(sparse_df, sparse_path)

sparse_seed <- DuckDBArraySeed(sparse_path,
                               datacol = "value",
                               keycols = list(i = 1:4, j = 1:3, k = 1:2))

# Extract as sparse array
sparse_block <- extract_sparse_array(sparse_seed, list(1:4, 1:3, 1))
sparse_block
```


## Construction

The *DuckDBArraySeed* constructor wraps *DuckDBTable* construction:

```{r DuckDBArraySeed-construct}
# From Parquet file
seed <- DuckDBArraySeed(array_path,
                        datacol = "value",
                        keycols = list(i = 1:4, j = 1:3, k = 1:2))

# From existing DuckDBTable
tbl <- DuckDBTable(array_path,
                   datacols = "value",
                   keycols = list(i = 1:4, j = 1:3, k = 1:2))
seed2 <- DuckDBArraySeed(tbl, datacol = "value",
                         keycols = list(i = 1:4, j = 1:3, k = 1:2))
```


## Type System

The seed inherits type detection from *DuckDBTable*:

```{r DuckDBArraySeed-type}
# Get array type
type(seed)

# Cast to different type
type(seed) <- "double"
type(seed)
```


# The DuckDBArray Class

## Class Structure

*DuckDBArray* extends *DelayedArray* and wraps a *DuckDBArraySeed*:

```{r DuckDBArray-class}
showClass("DuckDBArray")
```

This inheritance provides:
- *DelayedArray*: Lazy evaluation and block processing
- *DuckDBArraySeed*: DuckDB backend with SQL translation


## Construction

```{r DuckDBArray-construct}
# Create DuckDBArray from seed
arr <- DuckDBArray(seed)
arr

# Or directly from file
arr2 <- DuckDBArray(array_path,
                    datacol = "value",
                    keycols = list(i = 1:4, j = 1:3, k = 1:2))
```


## Array Operations

*DuckDBArray* supports the full array API:

### Subsetting

```{r DuckDBArray-subset}
# Subset by indices
arr[1:2, 1:2, 1]

# Subset by dimension
arr[, , 1]

# Drop dimensions
arr[1, , 1, drop = TRUE]
```

### Arithmetic

```{r DuckDBArray-arithmetic}
# Element-wise operations
arr * 2
arr + 10

# Operations between arrays
arr + arr
```

### Transformations

```{r DuckDBArray-transform}
# Permute dimensions
aperm(arr, c(2, 1, 3))
```


## Lazy Evaluation

Operations on *DuckDBArray* build a delayed operation tree:

```{r DuckDBArray-lazy}
# This doesn't execute immediately
result <- log1p(arr) * 2 + 1

# The seed is still DuckDBArraySeed
seed(result)

# Materialization happens on demand
as.array(result[1:2, 1:2, 1])
```


## Block Processing

For large arrays, operations can be processed in blocks:

```{r DuckDBArray-blocks, eval=FALSE}
library(DelayedArray)

# Set block size
setAutoBlockSize(1e8)  # 100 MB blocks

# Operations process blocks automatically
colSums(arr)  # Processes column-wise blocks
```


# The DuckDBMatrix Class

## Class Structure

*DuckDBMatrix* is a 2D specialization of *DuckDBArray*:

```{r DuckDBMatrix-class}
showClass("DuckDBMatrix")
```

### Validity Constraints

*DuckDBMatrix* requires exactly 2 dimensions in the seed.


## Construction

```{r DuckDBMatrix-construct}
# Create 2D data
matrix_df <- expand.grid(
    gene = paste0("GENE", 1:100),
    cell = paste0("CELL", 1:50)
)
matrix_df$count <- rpois(nrow(matrix_df), lambda = 5)

matrix_path <- tempfile(fileext = ".parquet")
arrow::write_parquet(matrix_df, matrix_path)

# Create DuckDBMatrix
mat <- DuckDBMatrix(matrix_path,
                    datacol = "count",
                    keycols = list(gene = paste0("GENE", 1:100),
                                   cell = paste0("CELL", 1:50)))
mat
```


### Row/Column Operations

```{r DuckDBMatrix-rowcol}
# Sums
rowSums(mat)
colSums(mat)

# Means
rowMeans(mat)
colMeans(mat)
```


# Matrix Statistics

## Optimized matrixStats Methods

`r Biocpkg("DuckDBArray")` provides SQL-optimized implementations of
`r CRANpkg("matrixStats")` functions:

```{r matrixStats}
library(MatrixGenerics)

# Row statistics
rowSums(mat)
rowMeans(mat)
rowVars(mat)
rowSds(mat)
rowMins(mat)
rowMaxs(mat)

# Column statistics
colSums(mat)
colMeans(mat)
colVars(mat)
```

These translate to efficient SQL aggregations:
```sql
-- rowSums becomes:
SELECT gene, SUM(count) FROM table GROUP BY gene

-- rowVars becomes:
SELECT gene, VAR_SAMP(count) FROM table GROUP BY gene
```


## Custom Statistics

The package includes specialized functions:

### rowDeviances()

Compute deviances for negative binomial models:

```{r rowDeviances}
# Deviance residuals for each gene
deviances <- rowDeviances(mat)
head(deviances)
```

### rowNnzs() / colNnzs()

Count non-zero entries:

```{r rowNnzs}
# Non-zero counts per gene
nnz_genes <- rowNnzs(mat)
head(nnz_genes)

# Non-zero counts per cell
nnz_cells <- colNnzs(mat)
head(nnz_cells)
```


# Sparse Arrays

## Sparse Representation

*DuckDBArray* efficiently handles sparse data:

```{r sparse-array}
# Create very sparse data (0.1% density)
set.seed(123)
n_nonzero <- 1000
sparse_df <- data.frame(
    i = sample(1:10000, n_nonzero),
    j = sample(1:10000, n_nonzero),
    value = rpois(n_nonzero, lambda = 10)
)

sparse_path <- tempfile(fileext = ".parquet")
arrow::write_parquet(sparse_df, sparse_path)

sparse_mat <- DuckDBMatrix(sparse_path,
                           datacol = "value",
                           keycols = list(i = 1:10000, j = 1:10000))

# Check sparsity
is_sparse(sparse_mat)

# Sparse operations are efficient
colSums(sparse_mat)  # Only processes non-zero entries
```


## Sparse Block Extraction

The `extract_sparse_array()` method returns sparse blocks:

```{r sparse-extract}
# Extract a sparse block
sparse_block <- extract_sparse_array(seed(sparse_mat), list(1:100, 1:100))
sparse_block

# Efficient storage
object.size(sparse_block)
```


## Coercion to Sparse Formats

```{r sparse-coerce}
# Coerce to COO_SparseArray
sparse_arr <- as(sparse_mat[1:100, 1:100], "COO_SparseArray")
sparse_arr

# Coerce to dgCMatrix (from Matrix package)
dgc <- as(sparse_mat[1:100, 1:100], "dgCMatrix")
class(dgc)
```


# Grid Partitioning

## The createDimTables() Function

For very large arrays, `createDimTables()` creates dimension lookup tables
that map indices to grid partitions:

```{r createDimTables}
# Create dimension tables for state.x77
library(S4Arrays)

state_grid <- RegularArrayGrid(dim(state.x77), c(10, 4))
dim_tables <- createDimTables(state.x77, grid = state_grid)

dim_tables
```

These tables enable:
1. **Efficient filtering**: Only read relevant partitions
2. **Parallel processing**: Process partitions independently
3. **Memory efficiency**: Load one partition at a time


## Using Dimension Tables

```{r dimtbls-usage}
# Create array with dimension tables
# (Usually done by writeParquet in BiocDuckDB)
arr_with_dims <- DuckDBArray(array_path,
                             datacol = "value",
                             keycols = list(i = 1:4, j = 1:3, k = 1:2),
                             dimtbls = dim_tables)

# Dimension tables attached to seed
dimtbls(seed(arr_with_dims))
```


# Advanced Topics

## Type Casting

Arrays can be cast to different types:

```{r type-casting}
# Original type
type(arr)

# Cast to integer
arr_int <- arr
type(arr_int) <- "integer"
type(arr_int)

# Cast to double
arr_dbl <- arr
type(arr_dbl) <- "double"
type(arr_dbl)
```


## Dimension Names

Arrays support dimension names:

```{r dimnames-ops}
# Set dimension names
dimnames(arr) <- list(
    paste0("R", 1:4),
    paste0("C", 1:3),
    paste0("L", 1:2)
)

arr

# Subset by names
arr[c("R1", "R2"), c("C1", "C2"), "L1"]
```


## Combining Arrays

```{r array-combine}
# Bind along existing dimension
arr1 <- arr[, , 1]
arr2 <- arr[, , 2]
combined <- rbind(arr1, arr2)
dim(combined)
```


# Integration with DelayedArray

## Delayed Operations

*DuckDBArray* participates in the DelayedArray ecosystem:

```{r delayed-ops}
library(DelayedArray)

# Combine with other DelayedArray backends
hdf5_path <- tempfile(fileext = ".h5")
library(HDF5Array)
h5_mat <- writeHDF5Array(matrix(rnorm(100), 10, 10), hdf5_path)

# Operations work across backends
duck_mat <- mat[1:10, 1:10]
# combined <- duck_mat + h5_mat  # Would work if dimensions match
```


## Block Processing

```{r block-processing}
# Custom block processing
process_blocks <- function(block) {
    # Custom operation on each block
    sum(block)
}

# Apply to blocks
grid <- defaultAutoGrid(mat)
blockApply(mat, process_blocks, grid = grid)
```


## Realization

Convert to other backends:

```{r realization, eval=FALSE}
# Realize as HDF5Array
h5_realized <- realize(mat, "HDF5Array")

# Realize as ordinary array
arr_realized <- as.array(mat)
```


# Use Cases

## Large Count Matrices

Single-cell RNA-seq count matrices are ideal for *DuckDBMatrix*:

```{r use-case-counts, eval=FALSE}
# Typical single-cell matrix: 30,000 genes × 100,000 cells
counts <- DuckDBMatrix("counts.parquet",
                       datacol = "count",
                       keycols = list(gene = gene_ids, cell = cell_ids))

# Efficient gene-level statistics
gene_means <- rowMeans(counts)
gene_vars <- rowVars(counts)

# Cell-level QC
cell_totals <- colSums(counts)
cell_detected <- colCounts(counts, value = 0)
```


## Sparse Genomic Data

Variant matrices or accessibility matrices are naturally sparse:

```{r use-case-sparse, eval=FALSE}
# scATAC-seq peak × cell matrix (99% sparse)
peaks <- DuckDBMatrix("peaks.parquet",
                      datacol = "accessibility",
                      keycols = list(peak = peak_ids, cell = cell_ids))

# Only non-zero entries stored in Parquet
# Efficient aggregations
peak_sums <- rowSums(peaks)  # Only processes non-zero values
```


## Multi-Dimensional Data

Higher-dimensional arrays for multi-condition experiments:

```{r use-case-multidim, eval=FALSE}
# Gene × Cell × Condition × Timepoint
multi_arr <- DuckDBArray("multi_cond.parquet",
                         datacol = "expression",
                         keycols = list(gene = genes,
                                       cell = cells,
                                       condition = conditions,
                                       time = timepoints))

# Subset to specific condition
cond1 <- multi_arr[, , "Treatment", ]
```


# Performance Considerations

## When to Use DuckDBArray

**Advantages**:
- Arrays larger than RAM
- Sparse data (< 10% density)
- Columnar aggregations (row/column sums, means)
- Parquet files already on disk
- Block-wise processing

**Limitations**:
- Random element access slower than in-memory
- Dense small arrays have overhead
- Some operations require materialization


## Optimization Tips

1. **Use sparse representation**: Store only non-zero entries
2. **Leverage grid partitioning**: For arrays > 1 GB
3. **Process by blocks**: Use DelayedArray block processing
4. **Filter early**: Subset before expensive operations
5. **Choose appropriate block size**: Balance memory and I/O


## Benchmarking

```{r benchmark, eval=FALSE}
# Compare with in-memory
library(microbenchmark)

# Small array (in-memory faster)
small_duck <- DuckDBMatrix(small_path, ...)
small_mem <- as.matrix(small_duck)

microbenchmark(
    duck = colSums(small_duck),
    memory = colSums(small_mem),
    times = 10
)

# Large array (DuckDB faster)
large_duck <- DuckDBMatrix(large_path, ...)
# large_mem <- as.matrix(large_duck)  # Won't fit in RAM!

system.time(colSums(large_duck))  # Efficient SQL aggregation
```


# Session Information

```{r sessioninfo}
sessionInfo()
```
