---
title: "Introduction to DuckDBArray"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r BiocStyle::doc_date()`; Modified: 6 July 2026"
package: DuckDBArray
vignette: >
  %\VignetteIndexEntry{1. Introduction to DuckDBArray}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      error = FALSE, warning = FALSE, message = FALSE)
```

# Introduction

Modern single-cell and genomic assays routinely produce matrices that do not fit
in memory. The 10x Genomics 1.3 million brain cell dataset, for example, is a
`27,998 x 1,306,127` matrix of UMI counts --- over 36 billion entries, or more
than 130 GB as an ordinary dense integer matrix. Even stored sparsely, datasets
like this are awkward to hold in an R session alongside the intermediate objects
an analysis creates.

`r Biocpkg("DelayedArray")` solves half of this problem: it lets an array live
*on disk* behind the ordinary R array API, so `dim()`, `[`, `rowSums()`, and
arithmetic all work without loading the data into memory. What remains is the
choice of **backend** --- the on-disk representation that actually stores and
serves the data. `r Biocpkg("HDF5Array")` provides an HDF5 backend and
`r Biocpkg("TileDBArray")` a TileDB one.

`r Biocpkg("DuckDBArray")` provides a backend built on **columnar Parquet files
queried through [DuckDB](https://duckdb.org)**. Rather than treating a count
matrix as a grid of blocks to be read and processed chunk by chunk, DuckDBArray
treats it as a *database table* and lets DuckDB's SQL engine do the work: a
`rowSums()` becomes a `SUM ... GROUP BY`, a `rowVars()` a `VAR_SAMP`, and the
query planner handles memory and multi-threading for you. As with any
`r Biocpkg("DelayedArray")` backend, **none of these operations load the matrix
into memory** --- they read only what they need from disk.

This vignette is a practical introduction: how to create a `DuckDBMatrix`, how
the familiar Bioconductor array operations behave on it, and when the DuckDB
backend is a good fit. Two companion vignettes go further:

- *Benchmarking DuckDBArray* compares the backend against in-memory,
  `r Biocpkg("HDF5Array")`, and `r Biocpkg("TileDBArray")` on real single-cell
  data.
- *Implementing the DuckDBArray backend* documents the seed contract and SQL
  translation for developers extending the package.

## Installation

```{r install, eval=FALSE}
if (!require("BiocManager"))
    install.packages("BiocManager")
BiocManager::install("DuckDBArray")
```

```{r load}
library(DuckDBArray)
```

# Quick start

A `DuckDBMatrix` is backed by a Parquet file in **coordinate (COO) form**: one
row per stored entry, with key columns giving the position and a data column
giving the value. For a small example we write such a file with
`r CRANpkg("arrow")`, storing a gene-by-cell count matrix as
`(gene, cell, count)` triples.

```{r quickstart-data}
library(arrow)

counts_df <- expand.grid(gene = paste0("Gene", 1:100),
                         cell = paste0("Cell", 1:20),
                         stringsAsFactors = FALSE)
set.seed(1L)
counts_df$count <- rpois(nrow(counts_df), lambda = 3)

parquet_path <- tempfile(fileext = ".parquet")
write_parquet(counts_df, parquet_path)
```

We construct the matrix by naming the data column and the key columns that index
each dimension. The key columns carry the dimension names and their order:

```{r quickstart-construct}
mat <- DuckDBMatrix(parquet_path,
                    datacol = "count",
                    keycols = list(gene = paste0("Gene", 1:100),
                                   cell = paste0("Cell", 1:20)))
mat
```

It looks and behaves like a matrix, but the data stays on disk:

```{r quickstart-ops}
dim(mat)
rowSums(mat)[1:5]
colMeans(mat)[1:5]
```

Nothing above read the whole matrix into memory: each summary is evaluated as a
DuckDB query over the Parquet file.

# Working with a DuckDBMatrix

## Writing a matrix to the DuckDB backend

In practice you usually start from an existing (possibly sparse) matrix. Use
`writeCoordArray()` to serialize it to Parquet, and `createDimTables()` to build
the dimension lookup tables that let DuckDB prune to the rows and columns a query
touches. Storing the matrix *transposed* puts features in columns, which
Parquet's columnar layout makes cheap to scan feature-by-feature.

```{r write-matrix}
library(Matrix)

m <- Matrix(rpois(200 * 50, lambda = 1), nrow = 200, ncol = 50, sparse = TRUE)
rownames(m) <- paste0("Gene", seq_len(nrow(m)))
colnames(m) <- paste0("Cell", seq_len(ncol(m)))

path <- tempfile()
writeCoordArray(m, path)
dimtbls <- createDimTables(m)

mat2 <- DuckDBMatrix(path, datacol = "value",
                     keycols = list(index1 = setNames(seq_len(nrow(m)), rownames(m)),
                                    index2 = setNames(seq_len(ncol(m)), colnames(m))),
                     dimtbls = dimtbls)
dim(mat2)
```

## Subsetting

Subsetting is delayed: it records the selection and is resolved only when values
are needed.

```{r subset}
sub <- mat[1:5, 1:4]
sub
as.matrix(sub)
```

## Matrix statistics

`r Biocpkg("DuckDBArray")` implements the `r Biocpkg("MatrixGenerics")` row/column
summaries as SQL aggregations, so the usual functions work directly:

```{r matrixstats}
library(MatrixGenerics)
rowSums(mat)[1:5]
rowVars(mat)[1:5]
colSds(mat)[1:5]
```

Two helpers are added for count-based feature selection --- `rowNnzs()` /
`colNnzs()` count non-zero entries (detection rate), and `rowDeviances()`
computes binomial/Poisson deviance for highly-variable-gene selection
([Townes et al. 2019](https://doi.org/10.1186/s13059-019-1861-6)):

```{r feature-selection}
rowNnzs(mat)[1:5]
rowDeviances(mat, family = "binomial")[1:5]
```

## Sparse data

Because only stored entries live in the Parquet file, sparse matrices stay
compact end to end. A `DuckDBMatrix` reports its sparsity and coerces to the
familiar sparse containers when you need them in memory:

```{r sparse}
is_sparse(mat2)
as(mat2[1:5, 1:4], "dgCMatrix")
```

## Lazy evaluation

A `DuckDBMatrix` *is* a `r Biocpkg("DelayedArray")`, so delayed operations
accumulate without touching the data, and materialize only on demand:

```{r lazy}
transformed <- log1p(mat) * 2
class(transformed)
as.matrix(transformed[1:3, 1:3])
```

# N-dimensional arrays

The same backend serves arrays of any dimension through `DuckDBArray()`; a
`DuckDBMatrix` is simply the two-dimensional case. Here the key columns index
three dimensions:

```{r nd-array}
arr_df <- expand.grid(i = 1:4, j = 1:3, k = 1:2)
arr_df$value <- rpois(nrow(arr_df), lambda = 2)
arr_path <- tempfile(fileext = ".parquet")
write_parquet(arr_df, arr_path)

arr <- DuckDBArray(arr_path, datacol = "value",
                   keycols = list(i = 1:4, j = 1:3, k = 1:2))
dim(arr)
arr[1:2, 1:2, 1]
```

# When to use DuckDBArray

DuckDBArray is a good fit when:

- the matrix is **larger than memory**, or you want to keep memory free for the
  rest of the analysis;
- the data is **sparse** (e.g. single-cell counts, scATAC-seq accessibility);
- your workload is dominated by **columnar aggregations** --- row/column
  summaries, grouped statistics, QC and feature selection;
- the data already lives on disk as **Parquet**, or benefits from a format that
  other tools (Python, Julia, cloud query engines) can read directly.

Other backends remain the better choice when data fits comfortably in memory (an
in-memory `dgCMatrix` is hard to beat for small data), or for workloads built
around dense linear algebra or heavy random element access.

For a head-to-head comparison with in-memory, `r Biocpkg("HDF5Array")`, and
`r Biocpkg("TileDBArray")` backends --- including where each one wins --- see the
*Benchmarking DuckDBArray* vignette. To understand or extend the backend itself,
see *Implementing the DuckDBArray backend*.

# Session information

```{r sessioninfo}
sessionInfo()
```
