---
title: "DuckDBArray: A DuckDB backend for DelayedArray"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
package: DuckDBArray
date: "Revised: December 18, 2025; Compiled: `r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{DuckDBArray: A DuckDB backend for DelayedArray}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 3
---

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

The `r Biocpkg("HDF5Array")` package has long been the workhorse for
out-of-memory single-cell analysis in Bioconductor. Its
[performance vignette](https://bioconductor.org/packages/release/bioc/vignettes/HDF5Array/inst/doc/HDF5Array_performance.html)
provides detailed benchmarks on the 10x Genomics 1.3 Million Brain Cell
Dataset, demonstrating how to tune block sizes and choose between sparse and
dense representations.

`r Biocpkg("DuckDBArray")` takes a different approach. Rather than treating
count matrices as arrays to be chunked and processed block-by-block, it treats
them as database tables to be queried. SQL databases have spent decades
optimizing exactly the operations we need for single-cell analysis: column
sums, row variances, and grouped aggregations.

This vignette demonstrates the **core DelayedArray backend** provided by DuckDBArray.
For higher-level integrations with scuttle and scran packages (QC metrics,
normalization, variance modeling, marker detection), see the BiocDuckDB package
vignette.

The results are striking. For basic matrix operations, DuckDB matches in-memory
performance while remaining disk-backed. For `rowDeviances`, DuckDB can actually
**outperform in-memory sparse matrices by 10x** by computing aggregations
directly via SQL.

# DuckDBArray capabilities

Before diving into benchmarks, let's survey what DuckDBArray provides. The key
insight is that **existing Bioconductor packages automatically use optimized
DuckDBArray methods**. When you call `rowSums()` on a DuckDBMatrix, it becomes
a SQL aggregation.

## Matrix statistics (MatrixGenerics)

All standard row/column summarization methods are implemented with SQL:

| Function | Description | SQL Operation |
|----------|-------------|---------------|
| `rowSums` / `colSums` | Sum across margin | `SUM()` |
| `rowMeans` / `colMeans` | Mean across margin | `AVG()` |
| `rowVars` / `colVars` | Variance | `VAR_SAMP()` |
| `rowSds` / `colSds` | Standard deviation | `STDDEV_SAMP()` |
| `rowMins` / `colMins` | Minimum value | `MIN()` |
| `rowMaxs` / `colMaxs` | Maximum value | `MAX()` |
| `rowCounts` / `colCounts` | Count specific value | `COUNT(*) WHERE` |

## Detection and feature selection

DuckDBArray adds specialized methods for single-cell analysis:

| Function | Description | Use Case |
|----------|-------------|----------|
| `rowNnzs` / `colNnzs` | Count non-zero elements | Detection rate, filtering |
| `rowDeviances` | Binomial/Poisson deviance | Feature selection (HVGs) |

These work on any matrix-like object (`matrix`, `dgCMatrix`, `DelayedMatrix`,
`DuckDBMatrix`) with consistent interfaces.

## Storage format

DuckDBArray stores sparse matrices in coordinate (COO) format using Parquet files:

```{r storage-example, eval=FALSE}
# Write a sparse matrix to DuckDB storage
library(DuckDBArray)
library(Matrix)

# Create a sparse matrix
mat <- Matrix(sample(0:10, 100*100, replace=TRUE, prob=c(0.9, rep(0.01, 10))),
              nrow=100, ncol=100, sparse=TRUE)

# Write to Parquet (transposed for efficient column access)
mat_t <- t(mat)
writeCoordArray(mat_t, "matrix.parquet")

# Create DuckDBMatrix with dimension tables for fast indexing
dimtbls <- createDimTables(mat_t)
mat_ddb <- DuckDBMatrix("matrix.parquet", datacol = "value",
                        keycols = list(
                            index2 = setNames(seq_len(ncol(mat_t)), colnames(mat_t)),
                            index1 = setNames(seq_len(nrow(mat_t)), rownames(mat_t))
                        ),
                        dimtbls = dimtbls)
```

# Setup

We configure block processing parameters for efficient handling of large
sparse matrices.

## Parallelism with disk-backed arrays

Disk-backed array backends face challenges with R's `MulticoreParam` that
prevent reliable parallel execution:
 
- **HDF5Array**: HDF5 uses file locking to prevent data corruption from
  concurrent writes. When R forks processes, each child tries to access
  the same file, triggering lock conflicts—even for read-only operations.
  
- **TileDBArray**: TileDB uses internal threading (Intel TBB) for parallel
  I/O. When R forks, child processes inherit the parent's thread state
  inconsistently, leading to hangs or crashes.

- **DuckDBMatrix**: No issues. DuckDB manages parallelism internally via
  its SQL query executor. Operations like `colSums`, `rowVars`, and
  `GROUP BY` aggregations automatically use multiple cores without
  requiring R-level forking.

For reliable benchmarking, we use `SerialParam` for R-level parallelism.
This ensures fair comparison across backends while allowing DuckDB to
leverage its internal parallel execution—a key architectural advantage.

```{r setup-blocks}
library(DelayedArray)
setAutoBlockSize(2^30)
setAutoBlockShape("scale")

# Use SerialParam for reliable execution across all disk-backed backends
# DuckDB still benefits from internal parallelism via its SQL engine
BPPARAM <- BiocParallel::SerialParam()
setAutoBPPARAM(BPPARAM)
```

# The test dataset

We use a subset of the 10x Genomics 1.3 Million Brain Cell Dataset, a
27,998 x 1,306,127 sparse matrix of UMI counts. This is the standard benchmark
used by `r Biocpkg("HDF5Array")` and is available via
`r Biocpkg("ExperimentHub")`.
```{r load-experimenthub}
library(ExperimentHub)
hub <- ExperimentHub()
hub["EH1039"]$description
```

```{r download-data, message=FALSE}
brain_path <- hub[["EH1039"]]
```

For this vignette build, we use a 12,500 cell subset to keep build times
short. The `inst/scripts` directory contains benchmarks run across a range of
dataset sizes (12,500 to 200,000 cells) and variable gene counts (1,000 to
2,000 genes).
```{r create-subset}
library(HDF5Array)
brain_full <- TENxMatrix(brain_path, group = "mm10")
dim(brain_full)

n_cells <- 12500
brain_hdf5 <- brain_full[, seq_len(n_cells)]
dim(brain_hdf5)

sparsity <- 1 - (sum(brain_hdf5 != 0) / length(brain_hdf5))
sprintf("%.1f%% of entries are zero", sparsity * 100)
```

# Setting up the backends

We compare four backends:

1. **In-memory sparse Matrix** (`dgCMatrix`): the baseline
2. **HDF5Array** (`TENxMatrix`): chunk-based HDF5 storage  
3. **TileDBArray**: tile-based storage
4. **DuckDBMatrix**: SQL-based columnar storage

## In-memory sparse matrix
```{r setup-memory}
library(Matrix)
time_to_mem <- system.time({
    brain_mem <- as(brain_hdf5, "dgCMatrix")
})
sprintf("Conversion to dgCMatrix: %.1f sec", time_to_mem["elapsed"])
format(object.size(brain_mem), units = "MB")
```

## HDF5Array
```{r setup-hdf5}
class(brain_hdf5)
is_sparse(brain_hdf5)
```

## TileDBArray
```{r setup-tiledb}
library(TileDBArray)
tiledb_path <- tempfile()
time_tiledb <- system.time({
    brain_tiledb <- writeTileDBArray(brain_mem, path = tiledb_path)
})
sprintf("Conversion to TileDBArray: %.1f sec", time_tiledb["elapsed"])
```

## DuckDBMatrix

We transpose the matrix before writing to Parquet so that features (genes)
become columns in the storage format. Since Parquet uses columnar compression,
this makes feature-wise access efficient: reading all values for a single gene
requires reading just one compressed column.
```{r setup-duckdb}
library(DuckDBArray)
duckdb_path <- tempfile()
time_duckdb <- system.time({
    brain_t <- t(brain_mem)
    writeCoordArray(brain_t, duckdb_path)
    dimtbls <- createDimTables(brain_t)
    brain_ddb <- DuckDBMatrix(duckdb_path, datacol = "value",
                              keycols = list(
                                  index2 = setNames(seq_len(ncol(brain_t)),
                                                    colnames(brain_t)),
                                  index1 = setNames(seq_len(nrow(brain_t)),
                                                    rownames(brain_t))
                              ),
                              dimtbls = dimtbls)
})
sprintf("Conversion to DuckDBMatrix: %.1f sec", time_duckdb["elapsed"])
dim(brain_ddb)
```

# Basic matrix operations

## Column sums
```{r bench-colsums}
time_colsums_mem <- system.time(cs_mem <- Matrix::colSums(brain_mem))
time_colsums_hdf5 <- system.time(cs_hdf5 <- colSums(brain_hdf5))
time_colsums_tiledb <- system.time(cs_tiledb <- colSums(brain_tiledb))
time_colsums_ddb <- system.time(cs_ddb <- colSums(brain_ddb))

all.equal(as.numeric(cs_mem), cs_hdf5)
all.equal(as.numeric(cs_mem), cs_tiledb)
all.equal(as.numeric(cs_mem), cs_ddb)

colsums_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "TileDBArray", "DuckDBMatrix"),
    colSums_sec = c(time_colsums_mem["elapsed"], time_colsums_hdf5["elapsed"],
                    time_colsums_tiledb["elapsed"], time_colsums_ddb["elapsed"])
)
knitr::kable(colsums_df)
```

## Row variances

Row variances are critical for identifying highly variable genes.
```{r bench-rowvars}
library(MatrixGenerics)

time_rowvars_mem <- system.time(rv_mem <- rowVars(brain_mem))
time_rowvars_hdf5 <- system.time(rv_hdf5 <- rowVars(brain_hdf5))
time_rowvars_tiledb <- system.time(rv_tiledb <- rowVars(brain_tiledb))
time_rowvars_ddb <- system.time(rv_ddb <- rowVars(brain_ddb))

all.equal(rv_mem, rv_hdf5)
all.equal(rv_mem, rv_tiledb)
all.equal(rv_mem, rv_ddb)

rowvars_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "TileDBArray", "DuckDBMatrix"),
    rowVars_sec = c(time_rowvars_mem["elapsed"], time_rowvars_hdf5["elapsed"],
                    time_rowvars_tiledb["elapsed"], time_rowvars_ddb["elapsed"])
)
knitr::kable(rowvars_df)
```

# Feature selection

Identifying highly variable genes (HVGs) is a critical step in single-cell
analysis. The traditional approach computes variance on normalized data, but
deviance-based selection using a multinomial model is specifically suited for
UMI count data ([Townes et al. 2019](https://doi.org/10.1186/s13059-019-1861-6)).

## Row deviances

DuckDBArray provides `rowDeviances()` with methods for `matrix`, `dgCMatrix`,
`DelayedMatrix`, and `DuckDBMatrix`. The DuckDB implementation translates the
deviance computation into SQL aggregations. We use binomial deviance, which
treats each UMI as a Bernoulli trial—the closest approximation to the
multinomial sampling model.

Note: The DuckDB implementation can actually be **faster than in-memory sparse
matrices** because SQL computes the deviance directly via aggregations,
avoiding the intermediate matrix operations required by the dgCMatrix method.
```{r bench-rowdeviances}
time_dev_mem <- system.time(dev_mem <- rowDeviances(brain_mem, family = "binomial"))
time_dev_hdf5 <- system.time(dev_hdf5 <- rowDeviances(brain_hdf5, family = "binomial"))
time_dev_tiledb <- system.time(dev_tiledb <- rowDeviances(brain_tiledb, family = "binomial"))
time_dev_ddb <- system.time(dev_ddb <- rowDeviances(brain_ddb, family = "binomial"))

all.equal(dev_mem, dev_hdf5)
all.equal(dev_mem, dev_tiledb)
all.equal(dev_mem, dev_ddb)

rowdev_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "TileDBArray", "DuckDBMatrix"),
    rowDeviances_sec = c(time_dev_mem["elapsed"], time_dev_hdf5["elapsed"],
                         time_dev_tiledb["elapsed"], time_dev_ddb["elapsed"])
)
knitr::kable(rowdev_df)
```

## Non-zero counts with rowNnzs

DuckDBArray's `rowNnzs()` efficiently counts non-zero elements. This is useful
for computing detection rates and filtering genes by expression frequency.

Note: TileDBArray currently lacks a `rowCounts` method for its sparse block
type (`dgTMatrix`), so we skip it in this benchmark.
```{r bench-rownnzs}
time_nnz_mem <- system.time(nnz_mem <- rowNnzs(brain_mem))
time_nnz_hdf5 <- system.time(nnz_hdf5 <- rowNnzs(brain_hdf5))
time_nnz_tiledb <- NA
time_nnz_ddb <- system.time(nnz_ddb <- rowNnzs(brain_ddb))

all.equal(as.integer(nnz_mem), as.integer(nnz_hdf5))
all.equal(as.integer(nnz_mem), as.integer(nnz_ddb))

nnz_df <- data.frame(
    Backend = c("In-memory", "HDF5Array", "TileDBArray", "DuckDBMatrix"),
    rowNnzs_sec = c(time_nnz_mem["elapsed"], time_nnz_hdf5["elapsed"],
                    NA, time_nnz_ddb["elapsed"])
)
knitr::kable(nnz_df)
```

# Summary

```{r summary-table}
summary_df <- data.frame(
    Operation = c("colSums", "rowVars", "rowDeviances", "rowNnzs"),
    InMemory = c(
        time_colsums_mem["elapsed"],
        time_rowvars_mem["elapsed"],
        time_dev_mem["elapsed"],
        time_nnz_mem["elapsed"]
    ),
    HDF5Array = c(
        time_colsums_hdf5["elapsed"],
        time_rowvars_hdf5["elapsed"],
        time_dev_hdf5["elapsed"],
        time_nnz_hdf5["elapsed"]
    ),
    TileDBArray = c(
        time_colsums_tiledb["elapsed"],
        time_rowvars_tiledb["elapsed"],
        time_dev_tiledb["elapsed"],
        NA
    ),
    DuckDB = c(
        time_colsums_ddb["elapsed"],
        time_rowvars_ddb["elapsed"],
        time_dev_ddb["elapsed"],
        time_nnz_ddb["elapsed"]
    )
)
summary_df$DuckDB_vs_HDF5 <- round(summary_df$HDF5Array / summary_df$DuckDB, 1)
knitr::kable(summary_df, digits = 1)
```

## Performance by category

The benchmarks reveal consistent patterns across operation types:

**Basic statistics** (`colSums`, `rowVars`): DuckDB matches in-memory
performance while remaining disk-backed. For `colSums`, DuckDB is **12x faster**
than HDF5Array (0.1s vs 1.0s); for `rowVars`, **16x faster** (0.4s vs 6.3s).
TileDBArray is generally slower than HDF5Array for most operations.

**Feature selection** (`rowDeviances`, `rowNnzs`): The `rowDeviances` result is
particularly striking: DuckDB is **10x faster than in-memory sparse matrix**
(0.5s vs 4.8s). This is because DuckDB's SQL implementation computes deviance
directly via aggregations, avoiding the intermediate matrix operations that the
dgCMatrix method requires. DuckDB is also **13x faster** than HDF5Array for this
operation (0.5s vs 6.4s). Detection counting (`rowNnzs`) shows DuckDB matching
in-memory performance while being **8.5x faster** than HDF5Array (0.3s vs 2.8s).

# Higher-level analysis

For QC metrics, normalization, pseudo-bulk aggregation, variance modeling, and
marker detection, see the `r Biocpkg("BiocDuckDB")` package. BiocDuckDB provides
optimized implementations of scuttle and scran methods that build on DuckDBArray's
efficient SQL backend.

Key BiocDuckDB capabilities:

- **QC metrics**: `perCellQCMetrics`, `perFeatureQCMetrics`
- **Normalization**: `normalizeCounts`, `librarySizeFactors`
- **Pseudo-bulk**: `summarizeAssayByGroup`
- **Variance modeling**: `modelGeneVar`, `modelGeneVarByPoisson`, `modelGeneCV2`
- **Marker detection**: `correlatePairs`, `findMarkers`, `scoreMarkers`

See the BiocDuckDB vignette "Motivation for a DuckDB backend for DelayedArray"
for comprehensive benchmarks of these methods.

# Running your own benchmarks

For more comprehensive benchmarks including normalization and PCA, DuckDBArray
provides scripts that you can run on your own hardware. The full benchmarks
were run across a matrix of dataset sizes:

- **Cell counts**: 12,500, 25,000, 50,000, 100,000, 200,000
- **Variable genes**: 1,000, 2,000

This matches the benchmark parameters used in the HDF5Array performance
vignette, enabling direct comparison between backends.

## Available scripts

The `inst/scripts` directory contains:

- `normalize_and_PCA.R`: Normalization and PCA benchmark
- `run_benchmarks.sh`: Full benchmark suite across cell counts

See `inst/scripts/README.md` for detailed usage instructions.

# Backend comparison

| Feature | dgCMatrix | HDF5Array | TileDBArray | DuckDBMatrix |
|---------|-----------|-----------|-------------|--------------|
| **Storage** | Memory | HDF5 file | TileDB array | Parquet file |
| **Memory usage** | Full data | Blocks only | Blocks only | Query results |
| **Block tuning** | N/A | Required | Required | Automatic |
| **Parallelism** | R-level | ❌ File locks | ❌ Thread conflicts | ✅ Internal |
| **SQL operations** | No | No | No | Yes |
| **Cloud storage** | No | Limited | S3/GCS | S3/GCS |
| **Cross-language** | R only | Python, R | Python, R | Universal |

## When to use each backend

**In-memory sparse Matrix** (`dgCMatrix`):
Use when data fits in memory. For most single-cell datasets under 500K cells,
this remains the fastest option.

**HDF5Array** (`TENxMatrix`):
Useful for native 10x Genomics format support. Requires block size tuning for
optimal performance. R-level parallel processing (`MulticoreParam`) is not
reliably supported due to HDF5 file locking.

**TileDBArray**:
Offers cloud storage integration (S3/GCS), array versioning, and time-travel
queries. Performance is generally slower than HDF5Array for most operations.
R-level parallel processing (`MulticoreParam`) conflicts with TileDB's internal
threading.

**DuckDBMatrix**:
Recommended for on-disk single-cell analysis when:

- Data is too large for memory
- Block size tuning is undesirable
- Parallel execution is important (DuckDB uses internal multi-threading)
- Cross-language data sharing is needed (Parquet is universal)
- You want near in-memory performance without loading data into RAM

## Key advantages of DuckDB

1. **Matches in-memory performance**: For basic operations like `colSums` and
   `rowVars`, DuckDB matches in-memory sparse matrix performance while
   remaining disk-backed.

2. **Sometimes faster than in-memory**: For complex operations like
   `rowDeviances`, DuckDB's SQL implementation is **10x faster than in-memory
   sparse matrices** by computing aggregations directly without intermediate
   matrix operations.

3. **No block size tuning**: The query planner handles memory management
   automatically, avoiding the 2-5x performance penalties from suboptimal
   block sizes that affect HDF5Array and TileDBArray.

4. **Built-in parallel execution**: DuckDB manages parallelism internally
   via its SQL query executor, automatically using multiple cores for
   aggregations and scans. Unlike HDF5Array and TileDBArray, DuckDB achieves
   parallel speedups without R-level forking.

5. **Interoperability**: Parquet files are readable by Python (`pyarrow`,
   `polars`), Julia, Rust, and cloud query services.

# Session information {-}

```{r}
sessionInfo()
```
